Wallace Session 2024-05-28
================

Please find below the R code history from your *Wallace* v2.1.2 session.

You can reproduce your session results by running this R Markdown file
in RStudio.

Each code block is called a “chunk”, and you can run them either
one-by-one or all at once by choosing an option in the “Run” menu at the
top-right corner of the “Source” pane in RStudio.

For more detailed information see <http://rmarkdown.rstudio.com>).

### Package installation

Wallace uses the following R packages that must be installed and loaded
before starting.

```{r}
library(spocc)
library(spThin)
library(dismo)
library(sf)
library(ENMeval)
library(wallace)
```

The *Wallace* session code .Rmd file is composed of a chain of module
functions that are internal to *Wallace*. Each of these functions
corresponds to a single module that the user ran during the session. To
see the internal code for these module functions, click on the links in
the .Rmd file. Users are encouraged to write custom code in the .Rmd
directly to modify their analysis, and even modify the module function
code to further customize. To see the source code for any module
function, just type its name into the R console and press Return.

```{r}
# example:
# just type the function name and press Return to see its source code
# paste this code into a new script to edit it
occs_queryDb
```

Your analyses are below.

------------------------------------------------------------------------

## Analysis for *Pseudopodoces humilis* (Ph)

User CSV path with occurrence data. If the CSV file is not in the
current workspace, change to the correct file path
(e.g. “/Users/darwin/Documents/occs/”).

```{r}
# NOTE: provide the folder path of the .csv file
occs_path <- ""
occs_path <- file.path(occs_path, "occurrence_thin5.csv")
# get a list of species occurrence data
userOccs_Ph <- occs_userOccs(
  txtPath = occs_path, 
  txtName = "occurrence_thin5.csv", 
  txtSep = ",", 
  txtDec = ".")
occs_Ph <- userOccs_Ph$Pseudopodoces_humilis$cleaned
```

### Obtain environmental data

Using user-specified variables.

```{r}
## Specify the directory with the environmental variables
dir_envs_Ph <- ""
envs_path <- file.path(dir_envs_Ph, c('bio02_local.tif', 'bio03_local.tif', 'bio07_local.tif', 'bio08_local.tif', 'bio11_local.tif', 'bio15_local.tif', 'bio18_local.tif', 'bio19_local.tif', 'elev_2.5m_local.tif'))
# Create environmental object 
envs_Ph <- envs_userEnvs(
  rasPath = envs_path,
  rasName = c('bio02_local.tif', 'bio03_local.tif', 'bio07_local.tif', 'bio08_local.tif', 'bio11_local.tif', 'bio15_local.tif', 'bio18_local.tif', 'bio19_local.tif', 'elev_2.5m_local.tif'),
  doBrick = FALSE)
occs_xy_Ph <- occs_Ph[c('longitude', 'latitude')]
occs_vals_Ph <- as.data.frame(raster::extract(envs_Ph, occs_xy_Ph, cellnumbers = TRUE))
# Remove duplicated same cell values
occs_Ph <- occs_Ph[!duplicated(occs_vals_Ph[, 1]), ]
occs_vals_Ph <- occs_vals_Ph[!duplicated(occs_vals_Ph[, 1]), -1]
# remove occurrence records with NA environmental values
occs_Ph <- occs_Ph[!(rowSums(is.na(occs_vals_Ph)) >= 1), ]
# also remove variable value rows with NA environmental values
occs_vals_Ph <- na.omit(occs_vals_Ph)
# add columns for env variable values for each occurrence record
occs_Ph <- cbind(occs_Ph, occs_vals_Ph)
```

### Process environmental data

Sampling of 10000 background points and corresponding environmental data
using a user provided background extent with a 0 degree buffer.

```{r}
# Load the user provided shapefile or csv file with the desired extent.
  ##User must input the path to shapefile or csv file and the file name 
# Define path
bgPath_Ph <- ""
bgExt_Ph <- penvs_userBgExtent(
  bgShp_path = paste0(bgPath_Ph, "niche_model_range", ".shp"),
  bgShp_name = paste0("niche_model_range", c(".shp", ".shx", ".dbf")),
  userBgBuf = 0,
  occs = occs_Ph)
# Mask environmental data to provided extent
bgMask_Ph <- penvs_bgMask(
  occs = occs_Ph,
  envs = envs_Ph,
  bgExt = bgExt_Ph)
# Sample background points from the provided area
bgSample_Ph <- penvs_bgSample(
  occs = occs_Ph,
  bgMask =  bgMask_Ph,
  bgPtsNum = 10000)
# Extract values of environmental layers for each background point
bgEnvsVals_Ph <- as.data.frame(raster::extract(bgMask_Ph,  bgSample_Ph))
##Add extracted values to background points table
bgEnvsVals_Ph <- cbind(scientific_name = paste0("bg_", "Pseudopodoces humilis"), bgSample_Ph,
                            occID = NA, year = NA, institution_code = NA, country = NA,
                            state_province = NA, locality = NA, elevation = NA,
                            record_type = NA, bgEnvsVals_Ph)
```

### Partition occurrence data

Partition occurrences and background points for model training and
validation using random k-fold, a non-spatial partition method.

```{r}
# R code to get partitioned data
groups_Ph <- part_partitionOccs(
  occs = occs_Ph ,
  bg =  bgSample_Ph, 
  method = "rand",
  kfolds = 5) 
```

### Build and Evaluate Niche Model

Generating a species distribution model using the maxent.jar algorithm
as implemented in ENMeval V2.0 (with clamping = TRUE). For tuning using
L, LQ, H feature classes and regularization multipliers in the 0.5, 3
range increasing by 0.5. Not using any categorical predictor variables.

```{r}
# Run maxent model for the selected species
model_Ph <- model_maxent(
  occs = occs_Ph,
  bg = bgEnvsVals_Ph,
  user.grp = groups_Ph, 
  bgMsk = bgMask_Ph,
  rms = c(0.5, 3), 
  rmsStep =  0.5,
  fcs = c('L', 'LQ', 'H'),
  clampSel = TRUE,
  algMaxent = "maxent.jar",
  parallel = TRUE,
  numCores = 4)
```

### Visualize

Generate a map of the Maxent generated model with no threshold

```{r}
# Select current model and obtain raster prediction
m_Ph <- model_Ph@models[["fc.H_rm.0.5"]]
predSel_Ph <- dismo::predict(
  m_Ph, bgMask_Ph,
  args = c(paste0("outputformat=", "cloglog"), 
           paste0("doclamp=", tolower(as.character(TRUE)))), 
  na.rm = TRUE)
#Get values of prediction
mapPredVals_Ph <- getRasterVals(predSel_Ph, "cloglog")
#Define colors and legend  
rasCols <- c("#2c7bb6", "#abd9e9", "#ffffbf", "#fdae61", "#d7191c")
legendPal <- colorNumeric(rev(rasCols), mapPredVals_Ph, na.color = 'transparent')
rasPal <- colorNumeric(rasCols, mapPredVals_Ph, na.color = 'transparent')
#Generate map
m <- leaflet() %>% addProviderTiles(providers$Esri.WorldTopoMap) 
m  %>%
  leaflet::addLegend("bottomright", pal = legendPal,
            title = "Predicted Suitability<br>(Training)",
            values = mapPredVals_Ph, layerId = "train",
            labFormat = reverseLabel(2, reverse_order = TRUE)) %>% 
  #add occurrence data
  addCircleMarkers(data = occs_Ph, lat = ~latitude, lng = ~longitude,
                   radius = 5, color = 'red', fill = TRUE, fillColor = "red",
                   fillOpacity = 0.2, weight = 2, popup = ~pop) %>% 
  ##Add model prediction
  addRasterImage(predSel_Ph, colors = rasPal, opacity = 0.7,
                 group = 'vis', layerId = 'mapPred', method = "ngb") %>%
 ##add background polygons
  addPolygons(data = bgExt_Ph,fill = FALSE,
              weight = 4, color = "blue", group = 'proj')
```

### Transfer model

Transferring the model to the same modelling area with no threshold
rule. Environmental layers for transferring are user provided.

```{r}
## Specify the directory with the environmental variables
dir_envsXf_Ph <- ""
envsXf_path <- file.path(dir_envsXf_Ph, c('bio02_local.tif', 'bio03_local.tif', 'bio07_local.tif', 'bio08_local.tif', 'bio11_local.tif', 'bio15_local.tif', 'bio18_local.tif', 'bio19_local.tif', 'elev_2.5m_local.tif'))
# Load user environmental variables
xferUserEnvs_Ph  <- envs_userEnvs(
  rasPath = envsXf_path,
  rasName = c('bio02_local.tif', 'bio03_local.tif', 'bio07_local.tif', 'bio08_local.tif', 'bio11_local.tif', 'bio15_local.tif', 'bio18_local.tif', 'bio19_local.tif', 'elev_2.5m_local.tif')
)
# Transfer model 
xfer_userEnvs_Ph <- xfer_userEnvs(
  evalOut = model_Ph,
  curModel= "fc.H_rm.0.5",
  envs = xferUserEnvs_Ph ,
  xfExt = bgExt_Ph,
  alg = "maxent.jar",
  outputType = "cloglog",
  clamp = TRUE
 )
# store the cropped variables of transfer
xferExt_Ph <- xfer_userEnvs_Ph$xferExt

###Make map of transfer
bb_Ph <-  bgExt_Ph@bbox
bbZoom <- polyZoom(bb_Ph[1, 1], bb_Ph[2, 1], bb_Ph[1, 2], 
                   bb_Ph[2, 2], fraction = 0.05)
mapXferVals_Ph <- getRasterVals(xfer_userEnvs_Ph$xferUser, "cloglog")
rasCols_Ph <- c("#2c7bb6", "#abd9e9", "#ffffbf", "#fdae61", "#d7191c")
# if no threshold specified
legendPal <- colorNumeric(rev(rasCols_Ph), mapXferVals_Ph, na.color = 'transparent')
rasPal_Ph <- colorNumeric(rasCols_Ph, mapXferVals_Ph, na.color = 'transparent')
m <- leaflet() %>% addProviderTiles(providers$Esri.WorldTopoMap) 
m %>%
  fitBounds(bbZoom[1], bbZoom[2], bbZoom[3], bbZoom[4]) %>%
  leaflet::addLegend("bottomright", pal = legendPal,
            title = "Predicted Suitability<br>(Transferred)",
            values = mapXferVals_Ph, layerId = 'xfer',
            labFormat = reverseLabel(2, reverse_order = TRUE)) %>%
# map model prediction raster and polygon of transfer
  clearMarkers() %>% clearShapes() %>% removeImage('xferRas') %>%
  addRasterImage(xfer_userEnvs_Ph$xferUser, colors = rasPal_Ph, opacity = 0.7,
                 layerId = 'xferRas', group = 'xfer', method = "ngb") %>%
 ##add polygon of transfer (same modeling area)
  addPolygons(data = bgExt_Ph, fill = FALSE,
              weight = 4, color = "blue", group = 'xfer')
```
